\Ch{Declarations}{Decl}
\Sec{Preamble}{Decl.Pre}
\p Declarations generally specify how names are to be interpreted. Declarations have the form
\begin{grammar}
  \define{declaration-seq}\br
  \textit{declaration}\br
  \textit{declaration-seq declaration}

  \define{declaration}\br
  \textit{name-declaration}\br
  \textit{special-declaration}\br
  \textit{empty-declaration}

  \define{name-declaration}\br
  \textit{variable-declaration}\br
  \textit{function-declaration}\br
  \textit{namespace-declaration}\br
  \textit{record-declaration}\br
  \textit{template-declaration}\br
  \textit{type-alias-declaration}\br
  ...

  \define{special-declaration}\br
  \textit{export-declaration-group}\br
  \textit{cbuffer-declaration-group}\br
  ...

  \define{empty-declaration} \terminal{;}

\end{grammar}

\Sec{Specifiers}{Decl.Spec}
\Sub{General}{Decl.Spec.General}
\p The specifiers that can be used in a declaration are
\begin{grammar}
  \define{decl-specifier}\br
  \textit{function-specifier}\br
  ...
\end{grammar}

\Sub{Function specifiers}{Decl.Spec.Fct}

\p A \textit{function-specifier} can be used only in a function declaration.

\begin{grammar}
  \define{function-specifier}\br
  \texttt{export}\br
\end{grammar}

\p The \texttt{export} specifier denotes that the function has program linkage (\ref{Basic.Linkage.Program}).

\p The \texttt{export} specifier cannot be used on functions directly or indirectly within an unnamed namespace.

\p Functions with program linkage can also be specified in \textit{export-declaration-group} (\ref{Decl.Export}).

\p If a function is declared with an \texttt{export} specifier then all redeclarations of the same function must also use the \texttt{export} specifier or be part of \textit{export-declaration-group} (\ref{Decl.Export}).

\Sec{Declarators}{Decl.Decl}

\begin{grammar}
  \define{init-declarator-list}\br
  \textit{init-declarator}\br
  \textit{init-declarator-list} \terminal{,} \textit{init-declarator}\br

  \define{init-declarator}\br
  \textit{declarator} \opt{initializer}\br

  \define{declarator}\br
  \textit{declarator-id} \opt{attribute-specifier-seq}\br
  \textit{declarator} \textit{parameters-and-qualifiers}\br
  \textit{declarator} \terminal{[} \opt{constant-expression} \terminal{]} \opt{attribute-specifier-seq}\br

  \define{parameters-and-qualifiers}\br
  \terminal{(} \textit{parameter-declaration-clause} \terminal{)} \opt{cv-qualifier-seq} \opt{attribute-specifier-seq}\br

  \define{cv-qualifier-seq}\br
  \textit{cv-qualifier} \opt{cv-qualifier-seq}\br

  \define{cv-qualifier}\br
  \terminal{const}\br

  \define{declarator-id}\br
  \textit{id-expression}\br

\end{grammar}

\Sub{Arrays}{Decl.arrays}
% This might need to nest deeper as we flesh out the sections.

\p A declaration \textit{T D} where \textit{D} is of the form

\begin{grammar}
\textit{D`} \terminal{[} \textit{N} \terminal{]} \opt{attribute-specifier-seq}\br
\end{grammar}

declares an array of \textit{N} \textit{T`}, where \textit{N} is the
\textit{array bound}, and \textit{T`} is the derived type of the declarator
\textit{D`} and is called the \textit{contained type} of the array. \textit{N}
shall be a constant expression converted to unsigned integer type and shall have
a value greater than zero.

\p A declaration \textit{T D} where \textit{D} is of the form

\begin{grammar}
\textit{D`} \terminal{[} \terminal{]} \opt{attribute-specifier-seq}\br
\end{grammar}

declares an array of unknown bound, except when the bound is implicitly derived
as described below. A declaration may only declare an array of unknown bound for
global declarations of resource objects (\ref{Resources}).

\p Array declarations may be nested to form multidimensional arrays. When
declaring a multidimensional array, only the first (outermost) dimension may
have an unknown bound; all inner dimensions must have known bounds.

\begin{note}
  For example the declaration \texttt{int A[3][4];} declares \texttt{A} as an
  array of 3 elements, each of which is an array of 4 integers.

  Applying this example to the above grammar, \texttt{T} is \texttt{int},
  \texttt{D} is \texttt{A[3][4]}, \texttt{D'} is \texttt{A[3]}, \texttt{N} is 3,
  and \texttt{T'} is \texttt{int[4]}.
\end{note}

\p The contained type of an array of any form shall be a complete type.

\p A declarator with an omitted array bound is valid when declaring an object of
array type which is not a non-static data member and is initialized with an
initializer that follows the declarator. In this case the array bound is implied
by the number of elements provided by the initializer list following the rules
described in \ref{Decl.Init}.

\Sec{Initializers}{Decl.Init}

\p The process of initialization described in this section applies to all
initializers regardless of the context.

\begin{grammar}
  \define{initializer}\br
  brace-or-equal-initializer\br
  \terminal{(} expression-list \terminal{)}\br

  \define{brace-or-equal-initializer}\br
  \terminal{=} initializer-clause\br
  braced-init-list\br

  \define{initializer-clause}\br
  assignment-expression\br
  braced-init-list\br

  \define{braced-init-list}\br
  \terminal{\{} initializer-list \opt{\terminal{,}} \terminal{\}}\br
  \terminal{\{} \terminal{\}}\br

  \define{initializer-list}\br
  initializer-clause\br
  initializer-list \terminal{,} initializer-clause\br
\end{grammar}

\Sub{Aggregate Initialization}{Decl.Init.Agg}

\p An \textit{aggregate} is a vector, matrix, array, or class.

\p The subobjects of an aggregate have a defined order. For vectors and arrays
the order is increasing subscript order. For matrices it is increasing subscript
order with the subscript nesting such that in the notation
\texttt{Mat[M][N]}, the ordering is \(Mat[0][0]...Mat[0][N]...
Mat[M][0]...Mat[M][N]\). For classes the order is base class, followed by member
subobjects in declaration order.

\p A \textit{flattened ordering} of subobjects can be produced by performing a
depth-first traversal of the subobjects of an object following the defined
subobject ordering.

\p Each \textit{braced initializer list} is comprised of zero or more
\textit{initializer-clause} expressions, which is either another braced
initializer list or an expression which generates a value that either is or can
be implicitly converted to an rvalue. Each assignment-expression is an object,
which may be a scalar or aggregate type. A \textit{flattened initializer
sequence} is a sequence of expressions constructed by a depth-first traversal
over each assignment-expression in an initializer-list and performing a
depth-first traversal accessing each subobject of the assignment-expression.

\p If the target object is an array of unknown size, the object is assumed to
have \(m\) possible elements during parsing, where \(m>0\).

\p An initializer-list is a valid initializer if for each element
\(E_{n \bmod m}\) in the target object's flattened ordering there is a
corresponding expression \(E_n\) in the flattened initializer sequence, which
can be implicitly converted to the element's type. For arrays of unknown size,
the total number of expressions in the flattened initializer sequence must be a
multiple of the array's base element type.

\p An initializer-list is invalid if the flattened initializer sequence contains
more or fewer elements than the target object's flattened ordering, or if any
initializer \(I_n\) cannot be implicitly converted to the corresponding element
\(E_n\)'s type.

\Sec{Function Definitions}{Decl.Function}
\Sec{Attributes}{Decl.Attr}
\Sub{Semantic Annotations}{Decl.Attr.Semantic}
\Sub{Entry Attributes}{Decl.Attr.Entry}

\Sec{Export Declarations}{Decl.Export}

\p One or more functions with \textit{external linkage} can be also specified in the form of

\begin{grammar}
  \define{export-declaration-group}\br
  \texttt{export} \terminal{\{} \opt{function-declaration-seq} \terminal{\}}\br

  \define{function-declaration-seq}\br
  \textit{function-declaration} \opt{function-declaration-seq}
\end{grammar}

\p The \texttt{export} specifier denotes that every \textit{function-declaration} included in \textit{function-declaration-seq} has \textit{external linkage} (\ref{Basic.Linkage.External}).

\p The \textit{export-declaration-group} declaration cannot appear directly or indirectly within an unnamed namespace.

\p Functions with \textit{external linkage} can also be declared with an \texttt{export} specifier (\ref{Decl.Spec.Fct}).

\p If a function is part of an \textit{export-declaration-group} then all redeclarations of the same function must also be part on a \textit{export-declaration-group} or be declared with an \texttt{export} specifier (\ref{Decl.Spec.Fct}).
