\makeglossaries

\newacronym{hlsl}{HLSL}{High Level Shader Language}
\newacronym{dxc}{DXC}{DirectX Shader Compiler}
\newacronym{fxc}{FXC}{Legacy DirectX Shader Compiler}
\newacronym{c}{C}{C Programming Language}
\newacronym{cpp}{C++}{C++ Programming Language}
\newacronym{api}{API}{Application Programming Interface}
\newacronym{spmd}{SPMD}{Single Program Multiple Data}
\newacronym{simd}{SIMD}{Single Instruction Multiple Data}
\newacronym{simt}{SIMT}{Single Instruction Multiple Thread}

\newglossaryentry{spirv}
{
  name={SPIR-V},
  description={SPIR-V is a portable intermediate representation designed by the
  Khronos Group for representing GPU programs}
}

\newglossaryentry{dx}
{
  name={DirectX},
  description={DirectX is the multimedia \acrshort{api} introduced with Windows
  95.}
}

\newglossaryentry{isoC}
{
  name={ISO C standard (2011)},
  description={ISO/IEC 9899:2011: Standard for Programming Language C.}
}

\newglossaryentry{isoC23}
{
  name={ISO C standard (2023)},
  description={ISO/IEC 9899:2023: Standard for Programming Language C.}
}

\newglossaryentry{isoCPP}
{
  name={ISO C++ standard (2011)},
  description={ISO/IEC 14882:2011: Standard for Programming Language C++.}
}

\newglossaryentry{IEEE754}
{
  name={IEEE Standard 754},
  description={IEEE Standard For Floating Point Arithmetic}
}

\newglossaryentry{sm}
{ name={Shader Model}, description={Versioned hardware description included as
  part of the DirectX specification, which is used for code generation to a
  common set of features across a range of vendors.}
}

\newglossaryentry{lane}{
  name={Lane},
  description={The computation performed on a single element as described in the
  \acrshort{spmd} program. Also called: thread.}
}

\newglossaryentry{wave}{
  name={Wave},
  description={A group of \gls{lane}s which execute together. The number of
  \gls{lane}s in a Wave varies by hardware implementation. Also called: warp,
  SIMD-group, subgroup, or wavefront.}
}

\newglossaryentry{quad}{
  name={Quad},
  description={A group of four \gls{lane}s which form a cluster of adjacent
  computations in the data topology. Also called: quad-group or quad-wave. }
}

\newglossaryentry{threadgroup}{
  name={Thread Group},
  description={A group of \gls{lane}s which may be subdivided into one or more
  \gls{wave}s and comprise a larger computation. Also known as: group,
  workgroup, block or thread block.}
}

\newglossaryentry{dispatch}{
  name={Dispatch},
  description={A group of one or more \gls{threadgroup}s which comprise the
  largest unit of a shader execution. Also called: grid, compute space or index
  space.}
}
