\Ch{Lexical Conventions}{Lex}

\Sec{Unit of Translation}{Lex.Translation}

\p The text of \acrshort{hlsl} programs is collected in \textit{source} and
\textit{header} files. The distinction between source and header files is social
and not technical. An implementation will construct a \textit{translation unit}
from a single source file and any included source or header files referenced via
the \texttt{\#include} preprocessing directive conforming to the \gls{isoC}
preprocessor specification.

\p An implementation may implicitly include additional sources as required to
expose the \acrshort{hlsl} library functionality as defined in (\ref{Runtime}).

\Sec{Phases of Translation}{Lex.Phases}

\p \acrshort{hlsl} inherits the phases of translation from \gls{isoCPP}, with
minor alterations, specifically the removal of support for trigraph and digraph
sequences. Below is a description of the phases.

\begin{enumerate}
  \item Source files are characters that are mapped to the basic source
  character set in an implementation-defined manner.
  \item Any sequence of backslash (\texttt{\textbackslash}) immediately followed
  by a new line is deleted, resulting in splicing lines together.
  \item Tokenization occurs and comments are isolated. If a source file ends in
  a partial comment or preprocessor token the program is ill-formed and a
  diagnostic shall be issued. Each comment block shall be treated as a single
  white-space character.
  \item Preprocessing directives are executed, macros are expanded,
  \texttt{pragma} and other unary operator expressions are executed. Processing
  of \texttt{\#include} directives results in all preceding steps being executed
  on the resolved file, and can continue recursively. Finally all preprocessing
  directives are removed from the source.
  \item Character and string literal specifiers are converted into the
  appropriate character set for the execution environment.
  \item Adjacent string literal tokens are concatenated.
  \item White-space is no longer significant. Syntactic and semantic analysis
  occurs translating the whole translation unit into an implementation-defined
  representation.
  \item The translation unit is processed to determine required instantiations,
  the definitions of the required instantiations are located, and the
  translation and instantiation units are merged. The program is ill-formed if
  any required instantiation cannot be located or fails during instantiation.
  \item External references are resolved, library references linked, and all
  translation output is collected into a single output.
\end{enumerate}

\Sec{Character Sets}{Lex.CharSet}

\p The \textit{basic source character set} is a subset of the ASCII character set.
The table below lists the valid characters and their ASCII values:

\begin{center}
  \begin{tabular}{|| c | c | c ||}
    \hline
    Hex ASCII Value & Character Name & Glyph or C Escape Sequence \\
    \hline
    0x09 & Horizontal Tab & \texttt{\textbackslash t} \\
    0x0A & Line Feed & \texttt{\textbackslash n} \\
    0x0D & Carriage Return & \texttt{\textbackslash r} \\
    0x20 & Space & \\
    0x21 & Exclamation Mark & \texttt{!}\\
    0x22 & Quotation Mark & \texttt{"}\\
    0x23 & Number Sign & \texttt{\#}\\
    0x25 & Percent Sign & \texttt{\%}\\
    0x26 & Ampersand & \texttt{\&}\\
    0x27 & Apostrophe & \texttt{'}\\
    0x28 & Left Parenthesis & \texttt{(}\\
    0x29 & Right Parenthesis & \texttt{)}\\
    0x2A & Asterisk & \texttt{*}\\
    0x2B & Plus Sign & \texttt{+}\\
    0x2C & Comma & \texttt{,}\\
    0x2D & Hyphen-Minus & \texttt{-}\\
    0x2E & Full Stop & \texttt{.}\\
    0x2F & Solidus & \texttt{/}\\
    0x30 .. 0x39 & Digit Zero .. Nine & \texttt{0 1 2 3 4 5 6 7 8 9}\\
    0x3A & Colon & \texttt{:}\\
    0x3B & Semicolon & \texttt{;}\\
    0x3C & Less-than Sign & \texttt{<}\\
    0x3D & Equals Sign & \texttt{=}\\
    0x3E & Greater-than Sign & \texttt{>}\\
    0x3F & Question Mark & \texttt{?}\\
    0x41 .. 0x5A & Latin Capital Letter A .. Z &
        \texttt{A B C D E F G H I J K L M}\\
    & & \texttt{N O P Q R S T U V W X Y Z}\\
    0x5B & Left Square Bracket & \texttt{[}\\
    0x5C & Reverse Solidus & \texttt{\textbackslash}\\
    0x5D & Right Square Bracket & \texttt{[}\\
    0x5E & Circumflex Accent & \texttt{\textasciicircum}\\
    0x5F & Underscore & \texttt{\_}\\
    0x61 .. 0x7A & Latin Small Letter a .. z &
        \texttt{a b c d e f g h i j k l m}\\
    & & \texttt{n o p q r s t u v w x y z}\\
    0x7B & Left Curly Bracket & \texttt{\{}\\
    0x7C & Vertical Line & \texttt{|}\\
    0x7D & Right Curly Bracket & \texttt{\}}\\
    \hline
  \end{tabular}
\end{center}

\p An implementation may allow source files to be written in alternate
\textit{extended character sets} as long as that set is a superset of the
\textit{basic character set}. The \textit{translation character set} is an
\textit{extended character set} or the \textit{basic character set} as chosen by
the implementation.

\Sec{Preprocessing Tokens}{Lex.PPTokens}

\begin{grammar}
  \define{preprocessing-token}\br
  header-name\br
  identifier\br
  pp-number\br
  character-literal\br
  string-literal\br
  preprocessing-op-or-punc\br
  \textnormal{each non-whitespace character from the \textit{translation
  character set} that cannot be one of the above}
\end{grammar}\footnote{The preprocessor is inherited from C++ 11 with no
grammar extensions. It is specified here only for completeness.}

\p Each preprocessing token that is converted to a token shall have the lexical
form of a keyword, an identifier, a constant, a string literal or an operator or
punctuator.

\p Preprocessing tokens are the minimal lexical elements of the language during
translation phases 3 through 6 (\ref{Lex.Phases}). Preprocessing tokens can be
separated by whitespace in the form of comments, white space characters, or
both. White space may appear within a preprocessing token only as part of a
header name or between the quotation characters in a character constant or
string literal.

\p Header name preprocessing tokens are only recognized within
\texttt{\#include} preprocessing directives, \texttt{\_\_has\_include} expressions,
and implementation-defined locations within \texttt{\#pragma} directives. In
those contexts, a sequence of characters that could be either a header name or a
string literal is recognized as a header name.

\Sec{Tokens}{Lex.Tokens}

\begin{grammar}
  \define{token}\br
  identifier\br
  keyword\br
  literal\br
  operator-or-punctuator
\end{grammar}

\p There are five kinds of tokens: identifiers, keywords, literals, and
operators or punctuators. All whitespace characters and comments are ignored
except as they separate tokens.

\Sec{Comments}{Lex.Comments}

\p The characters \texttt{/*} start a comment which terminates with the
characters \texttt{*\textbackslash}. The characters \texttt{//} start a comment
which terminates at the next new line.

\Sec{Header Names}{Lex.Headers}

\begin{grammar}
  \define{header-name}\br
  \texttt{<} h-char-sequence \texttt{>}\br
  \texttt{"} h-char-sequence \texttt{"}

  \define{h-char-sequence}\br
  h-char\br
  h-char-sequence h-char

  \define{h-char}\br
  \textnormal{any character in the \textit{translation character set} except
  newline or \texttt{>}}

  \define{q-char-sequence}\br
  q-char\br
  q-char-sequence q-char

  \define{q-char}\br
  \textnormal{any character in the \textit{translation character set} except
  newline or \texttt{"}}
\end{grammar}

\p Character sequences in header names are mapped to header files or external
source file names in an implementation defined way.

\Sec{Preprocessing numbers}{Lex.PPNumber}

\begin{grammar}
  \define{pp-number}\br
  digit\br
  \texttt{.} digit\br
  pp-number \texttt{'} digit\br
  pp-number \texttt{'} non-digit\br
  pp-number \texttt{e} sign\br
  pp-number \texttt{E} sign\br
  pp-number \texttt{p} sign\br
  pp-number \texttt{P} sign\br
  pp-number \texttt{.}
\end{grammar}

\p Preprocessing numbers begin with a digit or period (\texttt{.}), and may be
followed by valid identifier characters and floating point literal suffixes
(\texttt{e+}, \texttt{e-}, \texttt{E+}, \texttt{E-}, \texttt{p+}, \texttt{p-},
\texttt{P+}, and \texttt{P-}). Preprocessing number tokens lexically include all
\textit{integer-literal} and \textit{floating-literal} tokens.

\p Preprocessing numbers do not have types or values. Types and values are
assigned to \textit{integer-literal}, \textit{floating-literal}, and
\textit{vector-literal} tokens on successful conversion from preprocessing
numbers.

\p A preprocessing number cannot end in a period (\texttt{.}) if the immediate
next token is a \textit{scalar-element-sequence} (\ref{Lex.Literal.Vector}). In
this situation the \textit{pp-number} token is truncated to end before the
period\footnote{This grammar formulation is not context-free and requires an
LL(2) parser.}.

%\Sec{Identifiers}{Lex.Ident}

%\Sec{Keywords}{Lex.Keywords}

%\Sec{Operators and Punctuators}{Lex.Operators}

\Sec{Literals}{Lex.Literals}

\Sub{Literal Classifications}{Lex.Literal.Kinds}

\begin{grammar}
  \define{literal}\br
  integer-literal\br
  character-literal\br
  floating-literal\br
  string-literal\br
  boolean-literal\br
  vector-literal
\end{grammar}

%\Sub{Integer Literals}{Lex.Literal.Int}

%\Sub{Character Literals}{Lex.Literal.Char}

\Sub{Floating-point Literals}{Lex.Literal.Float}

\begin{grammar}
  \define{floating-literal}\br
  fractional-constant \opt{exponent-part} \opt{floating-suffix}\br
  digit-sequence exponent-part \opt{floating-suffx}\br
  \define{fractional-constant}\br
  \opt{digit-sequence} \texttt{.} digit-sequence\br
  digit-sequence \texttt{.}\br
  \define{exponent-part}\br
  \texttt{e} \opt{sign} digit-sequence\br
  \texttt{E} \opt{sign} digit-sequence\br
  \define{sign} \textnormal{one of}\br
  \texttt{+} \texttt{-}
  \define{digit-sequence}\br
  digit\br
  digit-sequence digit
  \define{floating-suffix} \textnormal{one of}
  \texttt{h} \texttt{f} \texttt{l} \texttt{H} \texttt{F} \texttt{L}
\end{grammar}

\p A floating literal is written either as a \textit{fractional-constant} with
an optional \textit{exponent-part} and optional \textit{floating-suffix}, or as
an integer \textit{digit-sequence} with a required \textit{exponent-part} and
optional \textit{floating-suffix}.

\p The type of a floating literal is \texttt{float}, unless explicitly specified
by a suffix. The suffixes \texttt{h} and \texttt{H} specify \texttt{half}, the
suffixes \texttt{f} and \texttt{F} specify \texttt{float}, and the suffixes
\texttt{l} and \texttt{L} specify \texttt{double}.\footnote{This substantially
deviates from the implementations in \acrshort{fxc} and \acrshort{dxc}, but is
consistent with the
\href{https://learn.microsoft.com/en-us/windows/win32/direct3dhlsl/dx-graphics-hlsl-appendix-grammar\#floating-point-numbers}
{official documentation} and the behavior of GLSL. It is also
substantially simpler to implement and more regular than the existing
behaviors.} If a value specified in the source is not in the range of
representable values for its type, the program is ill-formed.

%\Sub{String Literals}{Lex.Literal.String}

%\Sub{Boolean Literals}{Lex.Literal.Bool}

\Sub{Vector Literals}{Lex.Literal.Vector}

\begin{grammar}
  \define{vector-literal}\br
  integer-literal \texttt{.} scalar-element-sequence\br
  floating-literal \texttt{.} scalar-element-sequence

  \define{scalar-element-sequence}\br
  scalar-element-sequence-x\br
  scalar-element-sequence-r

  \define{scalar-element-sequence-x}\br
  \texttt{x}\br
  scalar-element-sequence-x \texttt{x}

  \define{scalar-element-sequence-r}\br
  \texttt{r}\br
  scalar-element-sequence-r \texttt{r}
\end{grammar}

\p A \textit{vector-literal} is an \textit{integer-literal} or
\textit{floating-point} literal followed by a period (\texttt{.}) and a
\textit{scalar-element-sequence}.

\p A \textit{scalar-element-sequence} is a \textit{vector-swizzle-sequence}
where only the first vector element accessor is valid (\texttt{x} or
\texttt{r}). A \textit{scalar-element-sequence} is equivalent to a vector splat
conversion performed on the \textit{integer-literal} or
\textit{floating-literal} value (\ref{Conv.vsplat}).
